# Image Processing & Classfication Using Terra

## First trying the methods in rspatial.org/terra
```{r}
if (!file.exists("data/rs/samples.rds")) {
    download.file("https://biogeo.ucdavis.edu/data/rspatial/rs.zip", dest = "data/rs.zip")
    unzip("data/rs.zip", exdir="data")
}
```

```{r}
library(terra)
b2 <- rast('data/rs/LC08_044034_20170614_B2.tif')
# Green
b3 <- rast('data/rs/LC08_044034_20170614_B3.tif')
# Red
b4 <- rast('data/rs/LC08_044034_20170614_B4.tif')
# Near Infrared (NIR)
b5 <- rast('data/rs/LC08_044034_20170614_B5.tif')
```
```{r}
s = c(b5, b4, b3)
```

Create a multi-layer SpatRaster using the filenames

```{r}
filenames <- paste0('data/rs/LC08_044034_20170614_B', 1:11, ".tif")
landsat <- rast(filenames)
landsat
```

```{r}
par(mfrow=c(2,2))
```


```{r}
plot(b2, main="Blue", col=gray(0:100/100))
plot(b3, main = "Green", col = gray(0:100 / 100))
plot(b4, main = "Red", col = gray(0:100 / 100))
plot(b5, main = "NIR", col = gray(0:100 / 100))
```

### Create 3-band images as RGB

```{r}
landsatRGB <- c(b4, b3, b2)
plotRGB(landsatRGB, stretch = "lin")
```

```{r}
landsatFalseColor <- c(b5, b4, b3)
plotRGB(landsatFalseColor, stretch="lin")
```

### Subsetting

We won't need the last four bands, so we'll subset.  Then we'll rename them.

```{r}
landsat <- subset(landsat, 1:7)
names(landsat) <- c('ultra-blue', 'blue','green','red','NIR','SWIR1','SWIR2')
```

### Spatial subset or crop

```{r}
e <- ext(624387,635752,4200047,4210939)
landsatcrop <- crop(landsat,e)
lsRGB <- subset(landsatcrop,4:2)
plotRGB(lsRGB,stretch="lin")
```
```{r}
lsFalse <- subset(landsatcrop,5:3)
plotRGB(lsFalse,stretch='lin')
```

### Saving results

```{r}
writeRaster(landsatcrop, filename="cropped-landsat.tif",overwrite=T)
```

### Relation between bands

```{r}
pairs(landsatcrop[[1:2]], main="Ultra-blue versus Blue")
pairs(landsatcrop[[4:5]], main="Red vs NIR")
```

### Extract pixel values

```{r}
# polygons with LULC info
samp <- vect(readRDS('data/rs/lcsamples.rds'))
# generate 50 point samples from the polygons
set.seed(555)
ptsamp <- spatSample(samp, 50, 'regular')
# Use xy coordinates to extract spectral values 
df <- extract(landsat, ptsamp)
head(df)

```

### Spectral profiles

```{r}
ms <- aggregate(df[,-1], list(ptsamp$class), mean)
# instead of the first column, use row names
rownames(ms) <- ms[,1]
ms <- ms[,-1]
```

```{r}
LULCcolors <- c('darkred','yellow','burlywood','cyan','blue')
# create matrix from df
ms <- as.matrix(ms)
# First create an emply plot
plot(0, ylim=c(0,0.6), xlim=c(1,7),type='n',xlab="Bands",ylab="Reflectance")
# Add the different classes
for (i in 1:nrow(ms)){
  lines(ms[i,],type='l',lwd=3,lty=1,col=LULCcolors[i])
}
title(main="Spectral Signatures", font.main=2)
legend('topleft',rownames(ms),cex=0.8,col=LULCcolors,lty=1,lwd=3,bty='n')

```
## Basic mathematical operations

pixel-wise calculations like map algebra.  We'll use it to do indices.

```{r}
library(terra)
filenames <- paste0('data/rs/LC08_044034_20170614_B', 1:11, ".tif")
landsat <- rast(filenames)
landsatRGB <- landsat[[c(4,3,3)]]
landsatFCC <- landsat[[c(5,4,3)]]
```

### Vegetation indices

```{r}
vi <- function(img, k, i) {
  bk <- img[[k]]
  bi <- img[[i]]
  vi <- (bk - bi)/(bk + bi)
  return(vi)
}
```
```{r}
ndvi <- vi(landsat, 5, 4)
plot(ndvi, col=rev(terrain.colors(10)), main='NDVI')
```

### Histogram

```{r}
hist(ndvi, main='NDVI values', xlab='NDVI',ylab='Frequency',
     col='wheat',xlim=c(-0.5,1),breaks=30,xaxt='n')
axis(side=1,at=seq(-0.6,1,0.2), labels=seq(-0.6,1,0.2))
```

