---
title: "seabird_leftovers"
author: "Jerry Davis"
date: "12/26/2021"
output: html_document
---



```{r}
AOI <- vect(st_union(st_buffer(transJul2006, 10000)))
transJul2006v <- vect(transJul2006)
vorJul2006 <- crop(voronoi(transJul2006v), AOI)
r <- rast(vorJul2006, res=1000)
vr <- rasterize(vorJul2006, r, "avg_sal")
library(gstat)
d <- data.frame(geom(transJul2006v)[,c("x", "y")], as.data.frame(transJul2006v))
gs <- gstat(formula=avg_dep~1, locations=~x+y, data=d, nmax=5, set=list(idp = 0))
nn <- interpolate(r, gs, debug.level=0)
nnmsk <- mask(nn, vr)
plot(nnmsk,1)
```




### Start with libraries, function and and reading data

```{r message=F, results='hide'}
library(igisci); library(sf); library(terra); library(tidyverse); 
library(tmap); library(maptiles); library(readxl) 
RMSE <- function(observed, predicted) {
  sqrt(mean((predicted - observed)^2, na.rm=TRUE))
}
AOI <- vect(st_union(st_buffer(transJul2006, 10000)))
transJul2006v <- vect(transJul2006)
vorJul2006 <- crop(voronoi(transJul2006v), AOI)
```

```{r}
plot(AOI)
plotRGB(oceanBase)
polys(AOI)
points(transJul2006v)
```

### Temperature analysis


#### Voronoi and creation of raster template

We're going to develop a simple prediction using Voronoi polygons (@Voronoi), so we'll create these and see what they look like with the original data points also displayed. Note that we'll need to convert to S4 objects using `vect()`:

```{r fig.cap="Temperature by Voronoi polygons"}
plot(vorJul2006, "avg_tem")
```

Rasterize the voronoi polygons

```{r fig.cap="Temperature July 2006"}
r <- rast(vorJul2006, res=1000)  # builds a blank raster of the same extent
vr <- rasterize(vorJul2006, r, "avg_tem")
plot(vr)
plotRGB(oceanBase)
plot(vr, add=T, alpha=0.5)
```

```{r}
# n-fold cross validation
n <- 20
set.seed(5132015)
kf <- sample(1:n, size = nrow(transJul2006v), replace=TRUE) # get random numbers 
   # between 1 and 5 to same size as input data points SpatVector sierraV 
rmse <- rep(NA, n) # blank set of 5 NAs
for (k in 1:n) {
  test <- transJul2006v[kf == k, ]  # split into 5 sets of points, 4/5 going to training set,
  train <- transJul2006v[kf != k, ] #  and 1/5 to test, including all of the variables
  v <- voronoi(train)
  v
  test
  p <- terra::extract(v, test) # extract values from v (built from training data) at test locations
  rmse[k] <- RMSE(test$avg_tem, p$avg_tem)
}
rmse
mean(rmse)

```

**Relative performance** is derived with respect to null model we created above that simply compared the input values with the mean.

```{r}
null <- RMSE(mean(transJul2006v$avg_tem), transJul2006v$avg_tem)
null
perf <- 1 - (mean(rmse) / null)
round(perf, 3)

```

### Salinity


#### Voronoi and creation of raster template

We're going to develop a simple prediction using Voronoi polygons (@Voronoi), so we'll create these and see what they look like with the original data points also displayed. Note that we'll need to convert to S4 objects using `vect()`:

```{r fig.cap="Salinity mapped by Voronoi polygon"}
plot(vorJul2006, "avg_sal")
```

Rasterize the voronoi polygons

```{r}
vr <- rasterize(vorJul2006, r, "avg_sal")
plot(vr)
plotRGB(oceanBase)
plot(vr, add=T, alpha=0.5)
```

```{r}
# n-fold cross validation
n <- 20
set.seed(5132015)
kf <- sample(1:n, size = nrow(transJul2006v), replace=TRUE) # get random numbers 
   # between 1 and 5 to same size as input data points SpatVector sierraV 
rmse <- rep(NA, n) # blank set of 5 NAs
for (k in 1:n) {
  test <- transJul2006v[kf == k, ]  # split into 5 sets of points, 4/5 going to training set,
  train <- transJul2006v[kf != k, ] #  and 1/5 to test, including all of the variables
  v <- voronoi(train)
  v
  test
  p <- terra::extract(v, test) # extract values from v (built from training data) at test locations
  rmse[k] <- RMSE(test$avg_sal, p$avg_sal)
}
rmse
mean(rmse)
```

**Relative performance:**

```{r}
null <- RMSE(mean(transJul2006v$avg_sal), transJul2006v$avg_sal)
null
perf <- 1 - (mean(rmse) / null)
round(perf, 3)


```


### Fluorescence analysis

```{r fig.cap="Fluorescence mapped by Voronoi polygon"}
plot(vorJul2006, "avg_fluo")
```

Rasterize the voronoi polygons

```{r}
vr <- rasterize(vorJul2006, r, "avg_fluo")
plot(vr)
plotRGB(oceanBase)
plot(vr, add=T, alpha=0.5)
```

**Cross validation and relative performance**

```{r}
# n-fold cross validation
n <- 20
set.seed(5132015)
kf <- sample(1:n, size = nrow(transJul2006v), replace=TRUE) # get random numbers 
   # between 1 and 5 to same size as input data points SpatVector sierraV 
rmse <- rep(NA, n) # blank set of 5 NAs
for (k in 1:n) {
  test <- transJul2006v[kf == k, ]  # split into 5 sets of points, 4/5 going to training set,
  train <- transJul2006v[kf != k, ] #  and 1/5 to test, including all of the variables
  v <- voronoi(train)
  v
  test
  p <- terra::extract(v, test) # extract values from v (built from training data) at test locations
  rmse[k] <- RMSE(test$avg_fluo, p$avg_fluo)
}
null <- RMSE(mean(transJul2006v$avg_fluo), transJul2006v$avg_fluo)
perf <- 1 - (mean(rmse) / null)
round(perf, 3)

```


### Depth analysis

```{r fig.cap="Depth mapped by Voronoi polygon"}
plot(vorJul2006, "avg_dep")
```

Rasterize the voronoi polygons

```{r}
r <- rast(vorJul2006, res=1000)  # builds a blank raster of the same extent
vr <- rasterize(vorJul2006, r, "avg_dep")
plot(vr)
plotRGB(oceanBase)
plot(vr, add=T, alpha=0.5)
```

**Cross validation and relative performance**

```{r}
# n-fold cross validation
n <- 5
set.seed(5132015)
kf <- sample(1:n, size = nrow(transJul2006v), replace=TRUE) # get random numbers 
   # between 1 and 5 to same size as input data points SpatVector sierraV 
rmse <- rep(NA, n) # blank set of 5 NAs
for (k in 1:n) {
  test <- transJul2006v[kf == k, ]  # split into 5 sets of points, 4/5 going to training set,
  train <- transJul2006v[kf != k, ] #  and 1/5 to test, including all of the variables
  v <- voronoi(train)
  v
  test
  p <- terra::extract(v, test) # extract values from v (built from training data) at test locations
  rmse[k] <- RMSE(test$avg_dep, p$avg_dep)
}
#rmse
#mean(rmse)
null <- RMSE(mean(transJul2006v$avg_dep), transJul2006v$avg_dep)
#null
perf <- 1 - (mean(rmse) / null)
round(perf, 3)
```


#### Nearest neighbor interpolation

The previous assignment of data to Voronoi polygons can be considered to be a nearest neighbor interpolation where only one neighbor is used, but we can instead use multiple neighbors. In this case we'll use up to 5 (`nmax=5`). Presumably setting `idp` to zero makes it a nearest neighbor, along with the `nmax` setting.

```{r fig.cap="Nearest neighbor interpolation of depth"}
library(gstat)
d <- data.frame(geom(transJul2006v)[,c("x", "y")], as.data.frame(transJul2006v))
head(d)
gs <- gstat(formula=avg_dep~1, locations=~x+y, data=d, nmax=5, set=list(idp = 0))
nn <- interpolate(r, gs, debug.level=0)
nnmsk <- mask(nn, vr)
plot(nnmsk,1)
```

#### IDW

The previous nearest neighbor interpolation is the same as an inverse-distance weighted interpolator with a limited nmax (default is Inf) and a power of zero.  We'll try a slightly higher power:  0.5.

```{r fig.cap="IDW interpolation, power=0.5"}
library(gstat)
gs <- gstat(formula=avg_dep~1, locations=~x+y, data=d, set=list(idp=0.5))
idw <- interpolate(r, gs, debug.level=0)
idwr <- mask(idw, vr)
plot(idwr, 1)
```


#### Trend surfaces/ polynomials

This seems to just require providing the appropriate formula and either going globally (defaults of `nmax=Inf` and default `idp=0`) or locally (`nmax=10` seems to work).

```{r fig.cap="Linear trend, depth"}
library(gstat)
gs <- gstat(formula=avg_dep~x+y, locations=~x+y, data=d)
trend1 <- interpolate(r, gs, debug.level=0)
trend1r <- mask(trend1, vr)
plot(trend1r, 1)
```

2nd and higher order trends don't work -- can't use x^2, x*y, etc., might work like interactions.

```{r fig.cap="2nd order trend, depth"}
library(gstat)
gs <- gstat(formula=avg_dep~x+y+x^2+y^2+x*y, locations=~x+y, data=d)
trend2 <- interpolate(r, gs, debug.level=0)
trend2r <- mask(trend2, vr)
#plot(trend2r, 1)
```

```{r fig.cap="3rd order trend, depth"}
library(gstat)
gs <- gstat(formula=avg_dep~x+y+x^2+y^2+x*y+x^3+y^3+x^2*y+x*y^2, locations=~x+y, data=d)
trend3 <- interpolate(r, gs, debug.level=0)
trend3r <- mask(trend3, vr)
#plot(trend3r, 1)
```

Local polynomial, 3rd order trend, similarly doesn't work

```{r fig.cap="local polynomial, 3rd order, depth"}
library(gstat)
gs <- gstat(formula=avg_dep~x+y+x^2+y^2+x*y+x^3+y^3+x^2*y+x*y^2, locations=~x+y, nmax=10, data=d)
local3 <- interpolate(r, gs, debug.level=0)
local3r <- mask(local3, vr)
#plot(local3r, 1)
```


#### Kriging 

```{r}
gs <- gstat(formula=avg_dep~1, locations=~x+y, data=d)
v <- variogram(gs)
v
plot(v)
```

```{r}
fve <- fit.variogram(v, vgm(psill=20000, model="Exp", range=50000))
plot(variogramLine(fve,50000), type="l", ylim=c(0,100000))
points(v[,2:3], pch=20, col='red')
```

```{r}
k <- gstat(formula=avg_dep~1, locations=~x+y, data=d, model=fve)
# predicted values
kp <- interpolate(r, k, debug.level=0)
ok <- mask(kp, vr)
names(ok) <- c('prediction', 'variance')
plot(ok)
```

### kriging for Salinity

```{r}
gs <- gstat(formula=avg_sal~1, locations=~x+y, data=d)
v <- variogram(gs)
v
plot(v)
```

```{r}
fve <- fit.variogram(v, vgm(psill=10, model="Exp", nugget=8, range=50000))
plot(variogramLine(fve,50000), type="l", ylim=c(0,20))
points(v[,2:3], pch=20, col='red')

#fve <- fit.variogram(v, vgm(psill=20000, model="Exp", range=50000))
#plot(variogramLine(fve,50000), type="l", ylim=c(0,100000))



```

### kriging for Temperature


```{r}
gs <- gstat(formula=avg_tem~1, locations=~x+y, data=d)
v <- variogram(gs)
v
plot(v)
```

```{r}
fve <- fit.variogram(v, vgm(psill=2.5, model="Exp", nugget=1.5, range=50000))
plot(variogramLine(fve,50000), type="l", ylim=c(0,5))
points(v[,2:3], pch=20, col='red')
```

### kriging for Fluorescence

```{r}
gs <- gstat(formula=avg_fluo~1, locations=~x+y, data=d)
v <- variogram(gs)
v
plot(v)
```

```{r}
fve <- fit.variogram(v, vgm(psill=0.12, model="Sph", range=50000))
plot(variogramLine(fve,50000), type="l", ylim=c(0,0.12))
points(v[,2:3], pch=20, col='red')
```

```{r fig.cap="Ordinary Kriging, Fluorescence"}
k <- gstat(formula=avg_fluo~1, locations=~x+y, data=dfilt, model=fve)
kp <- interpolate(r, k, debug.level=0)
ok <- mask(kp, vr)
names(ok) <- c('prediction', 'variance')
plot(ok)
```

## Michael Dorman methods using stars for raster

Trying the above, which relied on terra, with a method that relies on stars for raster.

Based on http://132.72.155.230:3838/r/spatial-interpolation-of-point-data.html

```{r}
library(sf)
transects <- st_read(ex("SFmarine/transects.shp"))
transJul2006 <- transects %>% filter(month==7 & year==2006)
Sanctuaries <- st_read(ex("SFmarine/Sanctuaries.shp"))
cordell_bank <- st_read(ex("SFmarine/cordell_bank.shp"))
isobath_200 <- st_read(ex("SFmarine/isobath_200.shp"))
sefi <- st_read(ex("SFmarine/sefi.shp"))

```

Build a raster template.

```{r}
library(stars)
AOI <- st_union(st_buffer(transJul2006, 10000))
AOIras <- st_rasterize(AOI, dx=1000, dy=1000)
r <- rast(vorJul2006, res=1000) 



vorJul2006 <- crop(voronoi(transJul2006), AOI)

```




## other stuff


```{r}
fve <- fit.variogram(v, vgm("Sph"))
plot(variogramLine(fve), type="l")
points(v[,2:3], pch=20, col='red')
```




IDW

```{r}
library(gstat); library(terra)
transJuly <- transects %>% filter(month==7)
gs <- gstat(formula=avg_tem~1, data=transJuly, set=list(idp=2))
idw <- interpolate(templateRas, gs)
vr <- rasterize(transJuly, templateRas, "avg_tem")
idwr <- mask(idw, vr)
plot(idwr, 1)
```


```{r}
gs <- gstat(formula=avg_tem~1, data=transJuly)
fve <- fit.variogram(v, vgm(psill=171, model="Exp", range=100000))
plot(variogramLine(fve), type="l", maxdist = 100000)#, ylim=c(5,17))
points(v[,2:3], pch=20, col='red')
fve <- fit.variogram(v, vgm("Sph"))
plot(variogramLine(fve), type="l")
points(v[,2:3], pch=20, col='red')
```

## glm model with poisson family

```{r}
library(tidyverse); library(sf); library(igisci)
transects <- st_read(ex("SFmarine/transects.shp"))
transJul2006 <- transects %>% filter(month==7 & year==2006 & avg_tem>0 & avg_sal>0 & avg_fluo>0)
summary(glm(bfal~avg_tem+avg_sal+avg_fluo+avg_dep+dist_land+dist_isla+dist_200m+dist_cord, data=transJul2006,family=poisson))
summary(transJul2006)

```

```{r}
tmap_mode("view")
tm_shape(transJuly) + tm_dots(col="avg_dep")
```

